---
title: "Pollution_France"
author: "Maëlys Rapenne, Loan Rottier, Nathan Vidal and Simon Simus"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(readr)
library(tidyverse)
library(sf)
library(mapview)
library(RColorBrewer)
library(geoR)
library(gstat)
library(ggplot2)
library(viridis)
library(rnaturalearth)
```

Source of the data: <https://object.infra.data.gouv.fr/browser/ineris-prod/lcsqa%2Fconcentrations-de-polluants-atmospheriques-reglementes%2Ftemps-reel%2F2026%2F>

# Import the data

```{r}
data <- read_delim("Export Moy. journalière - O₃ - 2026-02-19.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE)
```

Keep only metropolitan

```{r}
data <- data %>% filter(Latitude > 40) %>% filter(Longitude < 10) %>% filter(Longitude > -10)
data_sf <- st_as_sf(data, coords = c("Longitude", "Latitude"),crs = 4326)
```

# EDA

```{r}
mapview(data_sf, 
        zcol = "valeur brute",
        col.regions = brewer.pal(9, "Reds"),
        layer.name = "Niveaux de Pollution",
        legend = TRUE)
```

```{r}

france_map <- ne_countries(scale = "medium", country = "France", returnclass = "sf")
data_sf <- st_transform(data_sf, st_crs(france_map))
ggplot(data_sf) + 
  geom_sf(data = france_map, fill = "gray95", color = "black", linewidth = 0.3) + 
  geom_sf(aes(color = `valeur brute`), size = 2) + 
  coord_sf(xlim = c(-5.2, 9.6), ylim = c(41.3, 51.1), expand = FALSE) +
  scale_color_gradient(low = "blue", high = "orange") +
  theme_bw()
```

The ozone concentration seems to be lower on the north of the France.

# Model

## Variogram

```{r}
plot(variog(coords = st_coordinates(data_sf), data = data_sf$`valeur brute`,
            option = "cloud", max.dist = 400))
```

```{r}
coords <- st_coordinates(data_sf)
valeurs <- data_sf$`valeur brute`
my_geodata <- as.geodata(cbind(coords, valeurs))
# trend = "cte" because we assume the mean to be constant over the space
v <- variog(my_geodata, max.dist = 20, trend="cte") 
plot(v, main = "Ozone concentration variogram", 
     xlab = "Distance (in degrees)", 
     ylab = "Semi-variance")
```

The nugget is high, around 40, meaning that the variation between two close stations is slightly high. The total variance can be seen where the curve forms a "plateau", then the total variance, or the sill, is around 400. So we deduce a range, around 7.

```{r}
esti_nugget <- 40
esti_sill <- 400
esti_range <- 8
```

Interprétation ?

-   **Le Nugget (Effet de pépite) :** L'intersection avec l'axe Y à l'origine. Si elle est haute, cela signifie qu'il y a beaucoup de micro-variations (mesures très différentes entre deux stations pourtant proches).

-   **Le Sill (Palier) :** Le niveau où la courbe stagne. C'est la variance totale de votre échantillon.

-   **Le Range (Portée) :** La distance (sur l'axe X) à laquelle la courbe atteint le palier. **C'est l'information capitale pour vous :** au-delà de cette distance, la pollution d'un point n'est plus corrélée à celle d'un autre.

## Kriging

```{r}
vc <- variogram(data_sf$`valeur brute` ~ 1, data_sf, cloud = TRUE)
v <- variogram(data_sf$`valeur brute` ~ 1, data_sf)
plot(vc)
```

We test the Gaussian model.

```{r}
vinitial <- vgm(psill = (esti_sill - esti_nugget), model = "Gau",
                range = esti_range, nugget = esti_nugget)
v_fit <- fit.variogram(v, model = vinitial)
plot(v, v_fit, cutoff = 1000)
```

```{r}
print(v_fit)
```

Now, we try with a spherical model.

```{r}
vinitial <- vgm(psill = (esti_sill - esti_nugget), model = "Sph",
                range = esti_range, nugget = esti_nugget)
v_fit_sph <- fit.variogram(v, model = vinitial)
```

Problem of singular model, we try with the coefficient from the gaussian model.

```{r}
esti_nugget <- v_fit$psill[1]
esti_sill <- v_fit$psill[2]
esti_range <- v_fit$range[2]
```

```{r}
vinitial <- vgm(psill = (esti_sill - esti_nugget), model = "Sph",
                range = esti_range, nugget = esti_nugget)
v_fit_sph <- fit.variogram(v, model = vinitial)
plot(v, v_fit_sph, cutoff = 1000)
```

Better fit

### Kriging

Grid creation

```{r}
france <- ne_countries(scale = 50, country = "France", returnclass = "sf")
france <- st_crop(france, xmin = -5, xmax = 10, ymin = 41, ymax = 52)
france_grid <- st_make_grid(france, cellsize = 0.1, what = "centers")
france_grid <- st_intersection(france_grid, france)
france_grid_sf <- st_sf(geometry = france_grid)
plot(france_grid_sf)
```

```{r}
k <- gstat(formula = data_sf$`valeur brute` ~ 1, data = data_sf, model = v_fit_sph)
kpred <- predict(k, france_grid_sf)
```

```{r}
kpred
```

```{r}
ggplot() + geom_sf(data = kpred, aes(color = var1.pred)) +
  geom_sf(data = data_sf) +
  scale_color_viridis(name = "Ozone concentration - prediction") + theme_bw()
```

```{r}
ggplot() + geom_sf(data = kpred, aes(color = var1.var)) +
  geom_sf(data = data_sf) +
  scale_color_viridis(name = "Ozone concentration - variance") + theme_bw()
```

High variance, especially at the border and on the "hole"
