---
title: "Spatial Analysis of Ozone Pollution in Metropolitan France"
subtitle: "Geostatistical modelling and Kriging interpolation"
author: "Maëlys Rapenne, Loan Rottier, Nathan Vidal, Simon Simus"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
set.seed(1)
```

```{r}
library(readr)
library(tidyverse)
library(sf)
library(mapview)
library(RColorBrewer)
library(geoR)
library(gstat)
library(ggplot2)
library(viridis)
library(rnaturalearth)
library(sp)
library(knitr)
library(kableExtra)
```

------------------------------------------------------------------------

# Introduction

Air quality monitoring is a major public health challenge, particularly regarding ground-level ozone, a pollutant with significant respiratory impacts. Understanding its spatial distribution is essential for identifying exposure risks across the territory. However, while we have precise measurements from monitoring stations, we lack information for the vast majority of the country where no sensors are installed.

In this study, we analyze ozone concentration data provided by the LCSQA (Laboratoire Central de Surveillance de la Qualité de l'Air). Our dataset consists of the mean ozone concentrations recorded by each sensor across France for the specific day of February 19, 2026. Ozone is a pollutant whose concentration varies continuously over space, making it a good example of geostatistical data because it is a spatial continuous phenomenon observed at a finite set of locations.

\~Source of the [data:\\](data:\){.uri} <https://object.infra.data.gouv.fr/browser/ineris-prod/lcsqa%2Fconcentrations-de-polluants-atmospheriques-reglementes%2Ftemps-reel%2F2026%2F>\~

The objective is to use the Kriging geostatistical method to interpolate these point-based measurements. By leveraging the spatial autocorrelation of the data, we aim to build a continuous grid that predicts ozone levels in unsampled areas, providing a comprehensive overview of air quality on that day.

Methodology :

\- Spatial exploration of ozone concentrations (EDA).

\- Characterisation of spatial dependence through variogram analysis.

\- Spatial interpolation across France using Kriging. (Comparison with Inverse Distance Weighting (IDW) via cross-validation.)

# Theoretical framework

Air pollution, like ozone concentration, is a phenomenon that varies continuously across space, however it can only be measured at a finite set of locations. The field of geostatistics addresses this challenge: how to model, analyse, and predict spatially continuous phenomena from discrete observations.

Formally, we model the ozone field as a realisation of a random process $\{Z(s) : s \in D\}$, where $D \subset \mathbb{R}^2$ is the study domain (here, metropolitan France) and $s$ denotes a geographic location.

## Gaussian Random Fields

A Gaussian Random Field (GRF) is the theoretical model most commonly assumed for the spatial process $Z(\cdot)$. A GRF is a collection of random variables indexed by space, such that every finite subset of those variables follows a multivariate normal distribution. Under this assumption, the process is entirely described by its mean and its covariance function, which simplifies both the theoretical framework and the estimation:

-   $\mathbb{E}[Z(s)]$,
-   $Cov(s_i, s_j) = \text{Cov}(Z(s_i), Z(s_j))$.

To make estimation feasible with a limited number of observations, we impose two assumptions.

**Second-order stationarity** (also called weak stationarity). Under this condition, the process has a constant mean over the entire domain:

$$\mathbb{E}[Z(s)] = \mu, \quad \forall s \in D$$

and the covariance between any two locations depends only on the vector separating them, not on their absolute positions:

$$\text{Cov}(Z(s), Z(s+h)) = C(h), \quad \forall s \in D, \; \forall h \in \mathbb{R}^2$$

A further simplification is **isotropy**: the covariance depends only on the distance between locations, not the direction. When this assumption does not hold, the covariance structure must account for directional effects.

## The Covariance function and the Variogram

### Covariance function

Under stationarity and isotropy, the covariance between any two locations reduces to a function of their distance only, which we write :$$C(h) = \sigma^2  \text{Corr}(h) $$ where $\sigma^2$ is the marginal variance of the process and $\text{Corr}(h)$ is a positive definite correlation function decreasing with distance $h$.

Several parametric families exist for \$\\text{Corr}(h)\$, such as the Matérn or Gaussian families. In this study, we use the \*\*Spherical\*\* model, defined as: $$
\text{Corr}(h) =
\begin{cases}
1 - \frac{3h}{2\phi} + \frac{h^3}{2\phi^3} & \text{if } h \leq \phi \\[6pt]
0 & \text{if } h > \phi
\end{cases}
$$

Unlike models that approach the sill asymptotically, the Spherical model reaches exactly zero at the range \$\\phi\$, making this parameter directly interpretable. Near the origin, the curve is linear, which reflects a realistic degree of irregularity for a physical field like ozone concentration.

### Variogram

The variogram is used to characterise the spatial dependence structure of a geostatistical dataset. For a process satisfying stationarity and isotropy, the variogram $2\gamma(\cdot)$ is defined as:

$$2\gamma(h) = \text{Var}[Z(s+h) - Z(s)] = \mathbb{E}\left[(Z(s+h) - Z(s))^2\right]$$

The function $\gamma(\cdot)$ is the semivariogram.

The semivariogram is estimated from data using the empirical semivariogram:

$$\hat{\gamma}(h) = \frac{1}{2|N(h)|} \sum_{N(h)} \left(Z(s_i) - Z(s_j)\right)^2$$

where $N(h) = \{(s_i, s_j) : \|s_i - s_j\| = h\}$ is the set of observation pairs separated by distance $h$, and $|N(h)|$ is the number of such pairs.

When plotted against distance, the empirical semivariogram typically exhibits a characteristic shape described by three key parameters:

-   The nugget is the value of the semivariogram at distance zero
-   The sill is the plateau value reached by the semivariogram at large distances, corresponding to the total variance of the process.
-   The range is the distance at which the semivariogram reaches the sill, beyond which observations are considered spatially uncorrelated.

The empirical semivariogram is an essential exploratory tool: an increasing trend with distance indicates the presence of spatial correlation, while a flat semivariogram suggests spatial independence. It also serves as the basis for fitting a parametric variogram model, which is a necessary step before performing spatial interpolation via Kriging.

## Kriging

### Fitted variogram

We now want to predict the ozone concentration in places where sensors are not available. To perform this predictions, we will use Kriging, which is a spatial interpolation method to get predictions at unsampled locations.

While the previous section analyzed the sample variogram (empirical observations), we cannot use it directly for prediction. In fact, kriging requires a continuous mathematical function to calculate weights for any possible distance.

To fit the variogram, we use the Weighted Least Squares (WLS) method. This method estimates the parameters $\lambda$ (nugget, sill, range) by minimizing the weighted difference between the empirical semi-variance $\hat{\gamma}(h_j)$ and the theoretical model $\gamma(h_j; \lambda)$:

$$\sum_{j=1}^{k} w_j [\hat{\gamma}(h(j)) - \gamma(h(j); \lambda)]^2$$

We first try a gaussian model, using the estimated sill, range and nugget from the empirical variogram, and then a spherical model. When fitting the spherical model with the coefficients from the empirical variogram, the model can not converge (singular model). Thus, we use the coefficients from the gaussian variogram we fitted just before, and we get good results.

### Kriging

Now that we have fitted a variogram, we can perform spatial interpolation through Ordinary Kriging. Suppose that our data, observed, are $Z(s_1), \dots, Z(s_n)$. We want to predict the value of $Z$ at a point $s_0 \in D$. The ordinary Kriging estimator of $Z(s_0)$ is the estimator $$\hat{Z}(s_0) = \sum_{i=1}^{n} \lambda_i Z(s_i)$$This estimator is the Best Linear Unbiased Estimator since it minimizes the mean-squared prediction error : $$E[(\hat{Z}(s_0) - Z(s_0))^2]$$

subject to the constraint $\sum_{i=1}^n \lambda_i = 1$

The weights $\lambda_i$ come from the fitted variogram. This ensures that the model captures the spatial autocorrelation (nugget, sill, and range) of the data. Beyond simple distance, the estimator accounts for station clustering by reducing the weight of redundant sensors.

Now that we have the estimator, we need to build the points on which we are going to predict the ozone concentration. We thus build a grid of points in France. Using the `st_make_grid()` function, we discretize the study area into a regular spatial grid.

------------------------------------------------------------------------

# Data import and preprocessing

```{r}
data <- read_delim(
  "Export Moy. journalière - O₃ - 2026-02-19.csv",
  delim = ";", escape_double = FALSE, trim_ws = TRUE
)
```

We restrict the dataset to metropolitan France by filtering on geographic coordinates, removing overseas territories and erroneous records.

```{r}
data <- data %>%
  filter(Latitude  >  40,
         Longitude < 10,
         Longitude > -10)
```

```{r}
data_sf <- st_as_sf(data, coords = c("Longitude", "Latitude"), crs = 4326)
```

------------------------------------------------------------------------

# Exploratory Data Analysis (EDA)

## Distribution of ozone concentrations and anomalies

```{r}
summary(data_sf$`valeur brute`)
```

The ozone concentrations range from 0 to 98 µg/m3, a mean of 67 and a median close to 70, suggesting a symmetric distribution. The interquartile range (57 to79µg/m3 ) is relatively narrow, indicating that most stations record similar moderate levels.

The minimum value of 0 seem to be abnormal /anomaly.

```{r}
#count stations with value lower than 20
sum(data_sf$`valeur brute` < 25, na.rm = TRUE)
#there is only one station with a value of 0, which is likely an error or a station with no data.
#we decide to remove it from the dataset
data_sf <- data_sf %>% filter(`valeur brute` > 0)
```

## Interactive map

```{r}
mapview(data_sf,
        zcol       = "valeur brute",
        col.regions = brewer.pal(9, "Reds"),
        layer.name = "Concentration Ozone O3 (µg/m3)",
        legend     = TRUE)
```

## Static map

```{r}
france_map <- ne_countries(scale = "medium", country = "France", returnclass = "sf")
data_sf <- st_transform(data_sf, st_crs(france_map))
ggplot(data_sf) + 
  geom_sf(data = france_map, fill = "gray95", color = "black", linewidth = 0.3) + 
  geom_sf(aes(color = `valeur brute`), size = 2) + 
  coord_sf(xlim = c(-5.2, 9.6), ylim = c(41.3, 51.1), expand = FALSE) +
  scale_color_gradient(low = "blue", high = "orange", name = "O3 (µg/m3)") +
  labs(title = "Daily average ozone concentrations — 2026-02-19",
       caption = "Source: LCSQA / INERIS") +
  theme_bw()
```

Ozone concentrations are clearly higher in the south of France (Mediterranean coast, Rhône valley) and lower in the north - North East. This north-south gradient is consistent with higher solar radiation in the south, which promotes ozone formation.

Additionally, eastern France (Alsace, Rhone alpes) shows higher variability between neighbouring station. It might be due to influence of Vosges and Alps that can create microclimates.

## Distribution of ozone values

```{r}
ggplot(data_sf, aes(x = `valeur brute`)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(aes(xintercept = mean(`valeur brute`, na.rm = TRUE)),
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    title = "Distribution of O3 daily average concentrations",
    x     = "O3 concentration",
    y     = "Number of stations"
  ) +
  theme_bw()
```

> The distribution is roughly symmetric around the mean , with no extreme outliers. This is a favourable condition for applying Kriging, which assumes Gaussian random fields.

------------------------------------------------------------------------

# Variogram analysis

## Variogram cloud

The variogram cloud plots all pairwise squared differences against separation distance, without averaging. It helps detect outliers and assess the overall spatial dependence structure before smoothing.

We set the cutoff to 600 km, as beyond this distance the number of station pairs becomes too small to reliably estimate the semi-variance.

```{r}

vc <- variogram(data_sf$`valeur brute` ~ 1, data_sf, cloud = TRUE,cutoff = 500)
plot(vc,
     main = "Empirical variogram values",
     xlab = "Distance",
     ylab = "Semi-variance")
```

As expected, nearby stations tend to have lower semi-variance than distant ones, confirming spatial autocorrelation in ozone concentrations. The wide scatter at all distances reflects the local variability observed on the map. \## Sample (empirical) variogram

```{r}
v <- variogram(data_sf$`valeur brute` ~ 1, data_sf,cutoff = 500)
plot(v,
     main = "Average Empirical variogram values",
     xlab = "Distance",
     ylab = "Semi-variance")
```

The semi-variance is expressed in (µg/m³)², since it represents squared differences of ozone concentrations.

The semi-variogram increases from 40 µg²/m⁶ at short distances and seems to stabilise around 220-230 µg²/m⁶ beyond 400 km, suggesting a range of approximately 400 km. The nugget (40 µg²/m⁶) is small relative to the sill (230), meaning most of the variability is spatially structured. We set the cutoff to 500 km as beyond this distance the number of station pairs becomes too small to reliably estimate the semi-variance.

# Variogram model fitting

## Initial parameter estimates

```{r}
esti_nugget <- 40
esti_sill   <- 230
esti_range  <- 400
```

## Gaussian model

The Gaussian model assumes very smooth spatial variation (infinitely differentiable process)

```{r}
vinitial_gau <- vgm(
  psill  = esti_sill - esti_nugget,
  model  = "Gau",
  range  = esti_range,
  nugget = esti_nugget
)
v_fit_gau <- fit.variogram(v, model = vinitial_gau)
plot(v, v_fit_gau,
     main = "Fitted Gaussian variogram",
     xlab = "Distance (degrees)",
     ylab = "Semi-variance (µg²/m⁶)")
```

The Gaussian model fits the empirical variogram reasonably well up to 250 km, capturing the initial rise and the flattening around 150 µg²/m⁶. However, it underestimates the semi-variance at larger distances (350-500 km), where the empirical points continue to rise above the fitted curve.

```{r}
print(v_fit_gau)
```

The Gaussian model estimates a nugget of 41.4 µg²/m⁶, a sill of 158.1 µg²/m⁶ (41.4 + 116.7), and a range of 126 km, meaning that ozone concentrations are no longer spatially correlated beyond \~126 km.

## Spherical model

The Spherical model is the most widely used in geostatistics. Unlike the Gaussian model which approaches the sill asymptotically, the Spherical model reaches it exactly at the range distance, making the range parameter directly interpretable.

```{r}
vinitial_sph <- vgm(
  psill  = esti_sill - esti_nugget,
  model  = "Sph",
  range  = esti_range,
  nugget = esti_nugget,
)
v_fit_sph <- fit.variogram(v, model = vinitial_sph)
plot(v, v_fit_sph,
     main = "Fitted Spherical variogram",
     xlab = "Distance",
     ylab = "Semi-variance")
```

The Spherical model fits the empirical variogram well across the entire distance range, following the points closely from 0 to 500 km. However, since the range is estimated at 657 km (see after) beyond our cutoff of 500 km — the sill is never reached within the observed data, which explains why the curve is still rising at the end of the plot.

```{r}
v_fit_sph
```

The Spherical model estimates a nugget of 35.8 µg²/m⁶, a sill of 249.8 µg²/m⁶ (35.8 + 213.9), and a range of 657 km, meaning that ozone concentrations are no longer spatially correlated beyond 657 km.

## Model comparison

```{r}
gau_sse <- attr(v_fit_gau, "SSErr")
sph_sse <- attr(v_fit_sph, "SSErr")

comparison <- tibble(
  Model    = c("Gaussian", "Spherical"),
  Nugget   = c(v_fit_gau$psill[1], v_fit_sph$psill[1]),
  `Partial sill` = c(v_fit_gau$psill[2], v_fit_sph$psill[2]),
  Range    = c(v_fit_gau$range[2], v_fit_sph$range[2]),
  SSE      = c(gau_sse, sph_sse)
) %>% mutate(across(where(is.numeric), ~round(., 3)))

comparison
```

The Spherical model has a much lower SSE (160 vs 439), indicating a better fit to the empirical variogram. We therefore select the Spherical model for Kriging interpolation.

------------------------------------------------------------------------

# Kriging interpolation

## Prediction grid

We create a regular 0.1° grid covering metropolitan France, clipped to the country boundary.

```{r}
france <- ne_countries(scale = 50, country = "France", returnclass = "sf")
france <- st_crop(france, xmin = -5, xmax = 10, ymin = 41, ymax = 52)

france_grid    <- st_make_grid(france, cellsize = 0.1, what = "centers")
france_grid    <- st_intersection(france_grid, france)
france_grid_sf <- st_sf(geometry = france_grid)
plot(france_grid_sf)
```

## Ordinary Kriging predictions

```{r}
k     <- gstat(formula = data_sf$`valeur brute` ~ 1, data = data_sf, model = v_fit_sph)
kpred <- predict(k, france_grid_sf)
kpred
```

We fit an ordinary kriging model using the spherica variogram and predict ozone concentrations at each point of the France grid. The \~ 1 formula assumes a constant unknown mean over space.

The output contains 6431 grid points covering metropolitan France. For each point, Kriging returns a predicted ozone concentration (var1.pred) and an associated prediction variance (var1.var). For example, the first point located on the Mediterranean coast (9.09°E, 41.43°N) predicts a concentration of 81.8 µg/m³ with a variance of 98.7 µg²/m⁶.

## Kriging prediction map

```{r}
ggplot() + geom_sf(data = kpred, aes(color = var1.pred)) +
  geom_sf(data = data_sf) +
  scale_color_viridis(name = "Ozone concentration - prediction with Ordinary Kringing") + theme_bw()
```

The Kriging map produces a smooth continuous surface of ozone concentrations across France, consistent with the north-south gradient observed in the EDA. The highest concentrations (\>80 µg/m3) are predicted in the south-west and on the Mediterranean coast, while the lowest values (\<40 µg/m2) are found in the north-east. The smoothness of the surface reflects the large range estimated by the Spherical model (657 km), meaning that each prediction is influenced by stations up to 657 km away.

## Kriging variance (uncertainty) map

```{r}
ggplot() + geom_sf(data = kpred, aes(color = var1.var)) +
  geom_sf(data = data_sf) +
  scale_color_viridis(name = "Ozone concentration - variance") + theme_bw()
```

The variance map shows that prediction uncertainty is lowest in areas with a high density of monitoring stations, particularly near Paris and the Rhone valley. Conversely, variance is highest in areas with sparse coverage such as the south-west coast, and Corsica. This is a key property of Kriging: uncertainty is directly driven by the spatial distribution of observations, not by the predicted values themselves.
